---
title: "DA R PROGRAMMING EXAM"
author: "DA class"
date: "22/7/2024"
output: html_document
---

 ## IMPORT THE DATASET CALLED employees_attrition.csv file 
 ## INTO RSTUDIO
 
 ## EMPLOYEE ATTRITION MEANS EMPLOYEES ARE LEAVING THE COMPANY
 ## FOR ONE REASON OR THE OTHER.
 
 ## THE ATTRITION COLUMN HAS TWO POSSIBLE VALUES YES - FOR EMPLOYEES
 ## THAT LEFT THE COMPANY AND NO FOR THOSE THAT ARE  STILL AROUND.
 
 ## NOTE
 
 ## 1. RUN EVERY CODE CHUNCK YOU CREATE SO THAT THE OUTPUT WILL
 ## BE DISPLAYED BELOW THE CODE CHUNK.
 
 ## 2. SAVE YOUR RMARKDOWN FILE WITH YOUR NAME FOR IDENTIFICATION
 
 ## 3. USE DPLYR FUNCTIONS TO ANSWER THE FOLLOWING QUESTIONS.
 
 
 
```{r Q1}

# load the tidyverse package into this session
library("tidyverse")
```
 
```{r  Q2}
## import the given employees_attrition.csv file 

emp_attr <- read.csv(file.choose())

```

```{r Q3 }

## SHOW THE STRUCTURE OF THE DATAFRAME

str(emp_attr)

```


```{r Q4 }

## HOW MANY ROWS AND COLUMNS ARE THERE IN THE DATAFRAME

dim(emp_attr)

```

```{r Q5}
## SHOW THE NUMBER OF EMPLOYEES THAT LEFT THE COMPANY AND THE 
## NUMBER THAT ARE STILL WITH THE COMPANY
## HINT - USE THE group_by() function to categorise.


emp_attr %>%
  group_by(attrition) %>%
  summarise(count = n())

```


```{r Q6}
## SHOW THE NUMBER OF EMPLOYEES THAT LEFT THE COMPANY IN EACH 
## DEPARTMENT.

emp_attr %>%
  filter(attrition == "Yes") %>%
  group_by(department) %>%
  summarise(count = n())



```



```{r Q7}
## SHOW THE NUMBER OF MALE AND FEMALE EMPLOYEES THAT LEFT THE COMPANY
## AND THE NUMBER OF MALE AND FEMALE EMPLOYEES THAT ARE STILL WITH THE COMPANY
## HINT - USE THE group_by() function

emp_attr %>%
  group_by(gender,attrition) %>%
  summarise(count = n())
  


```




```{r Q8}

## WHAT IS THE AVERAGE NUMBER OF YEARS WORKED IN THE COMPANY BY THOSE THAT LEFT THE COMPANY.

emp_attr %>%
  filter(attrition == "Yes") %>%
  summarise(avg_years_worked = mean(YearsAtCompany))

```



```{r Q9}

## USING GGPLOT AND DPLYR FUNCTION, VISUALISE USING COLUMN CHART THE 5 JOB ROLES WITH THE HIGHEST NUMBER OF EMPLOYEES THAT LEFT.
## HINT - USE ggplot(aes(x =  val, y = val , fill = val)) +
## geom_col()

emp_attr %>%
  group_by(job_role) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(5) %>%
  ggplot(aes(x = job_role, y = count)) +
  geom_col()  +
  coord_flip()



```




```{r Q10}
## SHOW THE NUMBER OF EMPLOYEES THAT LEFT THE COMPANY BASED ON  THE isMarried column. 

emp_attr %>%
  filter(attrition == "Yes") %>%
  group_by(isMarried) %>%
  summarise(count = n())


```



```{r Q11 }
## SHOW THE AVERAGE MONTHLY INCOME EARNED BY THOSE WHO LEFT 
## THE COMPANY. BREAK THE INFORMATION DOWN BY DEPARTMENTS.
## SORT FROM THE HIGHEST AVERAGE INCOME TO THE LOWEST
emp_attr %>%
  filter(attrition == "Yes") %>%
  group_by(department) %>%
  summarise(avg_income = mean(monthly_income)) %>%
  arrange(-avg_income)


```


```{r Q12}

## SHOW THE AVERAGE MONTHLY INCOME EARNED BY THOSE WHO ARE STILL WITH THE COMPANY. BREAK THE INFORMATION DOWN BY DEPARTMENTS SORT 
## FROM HIGHEST AVERAGE INCOME TO THE LOWEST

emp_attr %>%
  filter(attrition == "No") %>%
  group_by(department) %>%
  summarise(avg_income = mean(monthly_income)) %>%
  arrange(-avg_income)

```



